---
title: "Homework 2"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
```

```{r}
knitr::opts_chunk$set(
        echo = TRUE,
        warning = FALSE,
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() +theme(legend.position = "bottom"))

options(
  ggplot2.continuous.color = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_color_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

# Part 1: Simulate using the SIR model and check basic model outputs
Read the SIR function and code provided in the script. Then run it with the following conditions and parameters.
N = 1e5
I0 = 10
S0 = N-I0
beta = 0.5
gamma = 0.3
times = seq(1, 100, by = 1)

## Question 1
Based on the simulation, how many people are susceptible on Day 50? How many people are susceptible at the end of simulation?

## Question 2
Based on the simulation, how many people are infectious on Day 50? How many people are infectious at the end of simulation?

## Question 3
Based on the simulation, how many people have been infected on Day 50? How many people have been infected by the end of simulation?

## Question 4
Plot the outputs I and S for each day during the simulated time period.


# Part 2: Simple analyses of model outputs
Convert the simulated numbers (#S, #I, #cumI etc.) to fraction relative to the population size

## Question 5
Based on the simulation, what is the population susceptibility (i.e. %S) at the end of simulation? What percentage of the population are infectious at the end of simulation? What percentage of the population have been infected by the end of simulation?

## Question 6
Plot the outputs %I and %S for each day during the simulated time period.


# Part 3: Simple analyses of model outputs
Aggregate the simulated daily outputs to weekly outputs. See model code in the script.

## Question 7
How many people are infected on Week 7?

## Question 8
Plot number of Susceptibles mid-week over time.

## Question 9
Plot number of infectious mid-week over time.

## Question 10
Plot number of New Infectious each week over time. 
