---
title: 'Lab 2: SIR Model'
subtitle: Script to test the SIR Model
date: "2024-01-24"
author: 
- Wan Yang (original code)
- Stephen Uong (tidy edit)
output:
  html_document:
    df_print: paged
---

NOTE: Please focus on the questions in the lab slides (i.e. pdf file)

-   Those are the core questions you need to answer.
-   Any additional questions in this R script are meant as guidelines.

If you haven't had it installed yet, run the following first:

```{r, eval = FALSE}
install.packages('pacman')
```

The pacman::p_load() function loads and installs packages if they are not yet installed.

```{r}
library(pacman)
pacman::p_load(
  deSolve,
  tidyverse,
  magrittr,
  # for animated plots
  gganimate,
  gifski
)
```

# A. TUTORIAL

## A1. PRE-LAB: TRY THE ODE SOLVER

SIMPLE ODE:

$\large\frac{dy}{dx} = x^2$

### Step 1: Code the ODE model (or function here)

```{r}
myfunc <- function(x, y, parms) {
  dy <- x^2
  dy %>% list()
}
```

### Step 2: Specify initial conditions/parameters

```{r}
# The 'time' step to integrate upon state <- c(y = 0)
xs <- seq(from = 0, to = 10, by = 0.2) 
state <- c(y = 0) # Create labeled vector with y = 0
```

### Step 3: Call the ode function to generate the simulation

```{r}
out <- deSolve::ode(y = state,
                    times = xs,
                    func = myfunc,
                    parms = NULL)
```

The analytic solution for this problem (i.e. true solution here)

```{r}
y.true <- (1/3)*xs^3
```

### Step 4: Check model outputs

#### Checking data frame

```{r}
# Clean ode output
out_df <- out %>% 
  tibble::as_tibble() %>% # Change format from matrix to a tibble data frame
  tibble::add_column(y.true = y.true) %>% # Add in y.true as a column 
  dplyr::mutate_all(as.numeric) # Make all columns numeric (from deSolve data type)

# tidyr::# To see the whole output - the out data frame here
out_df %>% View()  

# Print column names using the colnames function, so you can refer to them by name
out_df %>% colnames()

# Print part of the model output to see data structure
out_df %>% head(nrow = 3) # this prints the first 3 lines of the model output
out_df %>% tail(nrow = 6) # this prints the last 6 lines of the model output

# To get the first column of the output (time)
  # use column indexing 
out_df %>% dplyr::select(1)
  # use the column name
out_df %>% dplyr::select(time)

# Similarly to get the 2nd column of the output
  # use column indexing 
out_df %>% dplyr::select(2)
  # use the column name
out_df %>% dplyr::select(y)


# Now try checking the rows (e.g., print 3rd row of the output)
out_df %>% dplyr::slice(3)

# Now try specific row & column
out_df %>% dplyr::slice(3) %>% dplyr::select(2)       # row 3, column 2
out_df %>% dplyr::slice(3) %>% dplyr::select(time) # row 3, column time
out_df %>% dplyr::filter(time == 3) # row where time == 3
out_df %>% dplyr::filter(time >= 3 & time < 4) # rows where time >= 3 & < 4
out_df %>% dplyr::filter(y >= 10 & y < 11) # row where y >= 10 & < 11

```

#### Plotting values

```{r}
# Now plot to check if the ode is working
  # Plot y and y.true
out_df %>% 
  ggplot(aes(x = time)) +
  geom_line(aes(y = y.true, color = 'Analytic solution: y.true'), linetype = 'dotted', size = 2) +
  geom_line(aes(y = y, color = 'Numerical solution: ODE'), linetype = 'solid') +
  scale_color_manual(values = c('red','black')) + 
  labs(x = 'Time', y = 'y', color = 'Solution Type', linetype = 'Solution Type') +
  theme(legend.position = 'bottom')
  # Plot y only
out_df %>% 
  ggplot(aes(x = time)) +
  geom_line(aes(y = y), linetype = 'solid') +
  labs(x = 'Time', y = 'y', color = 'Solution Type') 

```

## A2. SIMULATION USING THE SIR MODEL

### Step 1: Code the SIR model (or function here)

```{r}
SIR <- function(t, state, parameters) {
  c(state, parameters) %>% # Concatenate state and parameter variables
  as.list %>%              # Coerce to list object (needed for with function)
  with({                 # Refer to variables by short name within list (e.g., beta vs. parameters$beta) 
    # Rate of change
    dS <- -beta*S*I/N
    dI <-  beta*S*I/N - gamma*I
    # Optional: also compute the cumulative number of infection
    dcumI <- beta*S*I/N
    # Return the rate of change as a list
    c(dS, dI, dcumI) %>% list()
  }) 
}

```

### Step 2: Specify initial conditions/parameters

```{r}
N <- 1e5 # Population size
I0 <- 10 # Initial No. of Infectious people
S0 <- N - I0 # Initial No. of Susceptible people
# Store the initial conditions I & S together  
state <- c(S = S0,        
           I = I0, 
           cumI = I0)
# Store the model parameters, unit: per day
parameters <- c(beta = 0.5, 
                gamma = 0.3) 
# Set simulation time steps: 1 to 100 days here
times <- seq(from = 1, to = 100, by = 1)  
```

### Step 3: Call the ode function to generate the simulation

```{r}
sim <- ode(y = state,
           times = times,
           func = SIR,
           parms = parameters)
```

### Step 4: Check model outputs

```{r}
sim_df <- sim %>% 
  tibble::as_tibble() %>% # Change format from matrix to a tibble data frame
  dplyr::mutate_all(as.numeric) # Make all columns numeric

sim_df %>% View()  # to see what's in the model output 'sim'

# Print part of the model simput to see data structure
sim_df %>% head(nrow = 3) # this prints the first 3 lines of the model output
sim_df %>% tail(nrow = 1) # this prints the last line of the model output

# Print column names using the colnames function, so you can refer to them by name
sim_df %>% colnames()

# To get the column for simulation
sim_df %>% dplyr::select(1) # use column indexing 
sim_df %>% dplyr::select(time) # use the column name

# get output on specific day using column name
sim_df %>% dplyr::filter(time == 10)


```

## A3. PRACTICE: CODE IT YOURSELF

### 3.1. Get the column for simulated number of Susceptibles

```{r}

```

### 3.2. Get the column for simulated number of Infectious:

```{r}

```

### 3.3. Visualize model outputs

Plot the following:

```{r}
# S over time

# I over time

# Compare S and I in one graph

# Compare S and cumI - think about the trend and why

```

# B. HOMEWORK QUESTIONS (CODE IT YOURSELF)

## B1. SET 1: Code It Yourself

### Question 1

```{r}
# [LQ1] What is the number of susceptibles on Day 50? 


# [LQ1] What is the number of susceptibles at the end of the simulation? 


```

### Question 2

```{r}
# [LQ2] What is the number of Infectious on Day 50? 


# [LQ2] What is the number of Infectious at the end of the simulation? 


```

### Question 3

```{r}
# [LQ3] Based on the simulation, how many people have been infected by Day 50? 


# [LQ3] How many people have been infected by the end of simulation? 


```

### Question 4

```{r}
# [LQ4] CODE IT YOURSELF - Plot model simulated number of Susceptibles for each day during the simulation period


# [LQ4] CODE IT YOURSELF - Plot model simulated number of Infectious for each day over time during the simulation period


```

## B2. Output Processing 1. Processing and Visualizing Model Outputs

### B2.1. Convert numbers to fractions relative to the population size (for easier comparison across populations of different sizes)

```{r}
sim_df_frac <- sim_df %>% 
  dplyr::mutate(s_frac = (S/N)*100, # %S
                i_frac = (I/N)*100, # %I
                cumi_frac = (cumI/N)*100) # % cumulative infections
```

### B2.2. Plot fraction susceptible over time

```{r}
# Show as percentage: 
sim_df_frac %>% 
  ggplot(aes(x = times, y = s_frac)) + 
  geom_line() +
  labs(x = 'Time (days)', y = 'Fraction Susceptible (%)')
```

## B3. SET 2: Code It Yourself

### Question 5

```{r}
# [LQ5] Based on the simulation, What is the population susceptibility (i.e. %S) at the end of simulation? What percentage of the population are infectious at the end of simulation?  What percentage of the population have been infected by the end of simulation?


```

### Question 6

```{r}
# [LQ6] Plot the outputs %I and %S for each day during the simulated time period.

```

## B4. Output Processing 2

```{r}
# Convert the daily outputs to weekly outputs
numWk <- times %>% length() %>% magrittr::divide_by(7) %>% ceiling()
tmstep <- 7
## Get the number of:
## - Susceptibles/Infectious for 4th day of each week (i.e. middle of the week) 
## - New Infections - these are cumulative numbers so need to get the increment (i.e., new infection) for each week
# Get index values
idx_day4 <- seq(4, length.out = numWk - 1, by  = tmstep)
idx_day1 <- seq(1, length.out = numWk    , by  = tmstep) # Keep 1 extra row to calculate diff
# Subset rows in data frames
sim_day4 <- sim_df %>% 
  dplyr::slice(idx_day4)
sim_day1_change <- sim_df %>% 
  dplyr::slice(idx_day1) %>% 
  dplyr::mutate(cumI_change = lead(cumI) - cumI) %>% 
  dplyr::filter(!is.na(cumI_change)) # drop last row with missing change value
# Create data frame with all values
out.weekly <- tibble(
  # Week column is just the number of week
  Week = 0:(numWk - 1), 
  # Number susceptibles - this is sort of a snapshot of the epidemic
    # use the initial susceptibility for the 1st entry
  Number_Susceptibles   = c(S0, sim_day4        %>% pull(S)),
  # Number infectious - this is the prevalence
    # use the initial # infectious for the 1st entry
  Number_Infectious     = c(I0, sim_day4        %>% pull(I)),
  # Number cumulative infections
    # Cumulative infections on the start of week 2, 3, ... (0 for 1st week)
  Number_Infections_New = c(0 , sim_day1_change %>% pull(cumI_change)))
out.weekly


```

## B5. SET 3: Code It Yourself

### Question 7

```{r}
# [Q7] How many people are infected on Week 7? (0.5pt)


```

### Question 8

```{r}
# [LQ8] Plot number of Susceptibles mid-week over time (0.5pt)


```

### Question 9

```{r}
# [LQ9] Plot number of Infectious mid-week over time (0.5pt)

```

### Question 10

```{r}
# [LQ10] Plot number of New Infections each week over time (0.5pt)


```

# C. Code for Demo

You can use the following code to see the change in S and I step by step in an animated plot

```{r}
### Specify initial variables
# Note: S0, I0, parameters, and N are still the same as above
numwk <- 15 
tmstep <- 7 # weekly increase

### Run simulation by week
for (wk in 1:numwk){
  # For first week, initialize objects and preliminary state
  if(wk == 1){
    times_wk <- list()
    sim_wk   <- list()
    state_wk <- list(c(S = S0, I = I0, cumI = I0))
  }
  # For each week, specify times
  times_wk[[wk]] <- seq(from = tmstep*(wk-1) + 1,
                        to   = tmstep* wk    + 1,
                        by   = 1)
  # For each week, run simlulation
  sim_wk[[wk]] <- ode(y     = state_wk[[wk]],
                      times = times_wk[[wk]],
                      func  = SIR,
                      parms = parameters) %>% 
                  dplyr::as_tibble() %>% 
                  dplyr::mutate_all(as.numeric)
  # For next week, save new state (so this state can be used in that week)
  state_wk[[wk + 1]] <- c(S    = sim_wk[[wk]] %>% dplyr::slice(n()) %>% dplyr::pull(S),
                          I    = sim_wk[[wk]] %>% dplyr::slice(n()) %>% dplyr::pull(I),
                          cumI = sim_wk[[wk]] %>% dplyr::slice(n()) %>% dplyr::pull(cumI))
}

### Clean data
sim_all <- sim_wk %>% 
  dplyr::bind_rows() %>% # Coerce list to data frame
  dplyr::distinct() %>% # Remove duplicate rows (same S/I/cumI at end/beginning of weeks)
  dplyr::mutate(wk = (time/7) %>% ceiling(),
                s_frac = (S/N)*100, # %S
                i_frac = (I/N)*100, # %I
                cumi_frac = (cumI/N)*100)

### Plot S, I, and cumI (N) over time
# Static plot
sim_plot <- sim_all %>% 
  tidyr::pivot_longer(cols = c(S, I, cumI), 
                      names_to = 'state',
                      values_to = 'n') %>% 
  ggplot(aes(x = time, y = n, color = state)) + 
  geom_point() +
  labs(x = 'Day', y = 'N', color = 'State') +
  scale_color_manual(values = c('violetred4','violetred2','steelblue2')) +
  expand_limits(x = 0, y = 0) +
  theme_bw()
sim_plot
# Dynamic/animated plot
sim_anim <- sim_plot + 
  gganimate::transition_states(wk) +
  labs(title = 'Week: {closest_state}') +
  shadow_mark(alpha = 0.3, size = 0.5) 
sim_anim %>% gganimate::animate(renderer = gifski_renderer())


### Plot S, I, and cumI (%) over time
# Same code as above but just use the s_frac, i_frac, and cumi_frac columns now
# Static plot
sim_plot_frac <- sim_all %>% 
  tidyr::pivot_longer(cols = c(s_frac, i_frac, cumi_frac), 
                      names_to = 'state',
                      values_to = 'n') %>% 
  ggplot(aes(x = time, y = n, color = state)) + 
  geom_point() +
  labs(x = 'Day', y = 'Percentage', color = 'State') +
  scale_color_manual(values = c('violetred4','violetred2','steelblue2')) +
  expand_limits(x = 0, y = 0) +
  theme_bw()
sim_plot_frac
# Dynamic/animated plot
sim_anim_frac <- sim_plot_frac + 
  gganimate::transition_states(wk) +
  labs(title = 'Week: {closest_state}') +
  shadow_mark(alpha = 0.3, size = 0.5) 
sim_anim_frac %>% 
  gganimate::animate(renderer = gifski_renderer())

```
