---
title: "SIR Simulations"
output: github_document
---

```{r setup, include=FALSE}
library(deSolve)
```

# Ordinary Differential Equations (ODE)

**General Routine**:    
  1. Code the ODE/model (or function here)    
  2. Specify the initial conditions/parameters    
  3. Call the "ode" function to solve the ODE/model and generate the simulation   
  4. Check/analyze model outputs
  
## Step 1: Code the ODE Model (or function here)
```{r}
myfunc = function(x, y, parms) {
  dy = x^2
  list(dy)
}
```

## Step 2: Specify Initial Conditions/Parameters
```{r}
xs = seq(0, 10, by = 0.2) #the 'time' step to integrate upon
state = c (y = 0);
```

## Step 3: Call the ODE Function to Generate the Simulation
```{r}
out = ode (y = state, times = xs, func = myfunc, parms = NULL)

#the analytical solution for this problem
y.true = 1/3 * xs^3
```

## Step 4: Check/Analyze Model Outputs
```{r}
print(out) # to view the entire output
colnames(out) #showing column names so you can refer to them by name

#plotting to check the ode is working
par(mar = c(3, 3, 1, 1), mgp = c(1.8, 0.5, 0), cex = 1.2)
plot(x = out[, 1], y = out[, 2], xlab = 'x', ylab = 'y', type = 'l')
points(x = xs, y = y.true, col = 'red', pch = 20)
legend('topleft', legend = c('Analytical solution: y.true', 'Numerical solution: ODE'), 
       col = c('red', 'black'), lty = c(NA, 1), pch = c(20, NA), bty = 'n')

#another method to check the ode is working
plot(x = out[, 'time'], y = out[, 'y'], xlab = 'x', ylab = 'y', type = 'l')
```

```{r}
#side by side comparison
par(mar = c(3, 3, 1, 1), mgp = c(1.8, 0.5, 0), cex = 1.2,
    mfrow = c(2, 1)) 
plot(x = out[, 1], y = out[, 2], xlab = 'x', ylab = 'y', type = 'l', 
     main = "Using column indices")  
points(x = xs, y = y.true, col = 'red', pch = 20)
legend('topleft', legend = c('Analytic solution: y.true', 'Numerical solution: ODE'),
       col = c('red', 'black'), lty = c(NA, 1), pch = c(20, NA), bty = 'n')


plot(x = out[, 'time'], y = out[, 'y'], xlab = 'x', ylab = 'y', type = 'l', 
     main = "Using column names")
points(x = xs, y = y.true, col = 'red', pch = 20)
legend('topleft', legend = c('Analytic solution: y.true', 'Numerical solution: ODE'),
       col = c('red', 'black'), lty = c(NA, 1), pch = c(20, NA), bty = 'n')


par(mar = c(3, 3, 1, 1), mgp = c(1.8, 0.5, 0), cex = 1.2,
    mfrow = c(1, 2)) 
plot(x = out[, 1], y = out[, 2], xlab = 'x', ylab = 'y', type = 'l', 
     main = "Using column indices")  
points(x = xs, y = y.true, col = 'red', pch = 20)
legend('topleft', legend = c('Analytic solution: y.true', 'Numerical solution: ODE'),
       col = c('red', 'black'), lty = c(NA, 1), pch = c(20, NA), bty = 'n')

plot(x = out[, 'time'], y = out[, 'y'], xlab = 'x', ylab = 'y', type = 'l', 
     main = "Using column names") 
points(x = xs, y = y.true, col = 'red', pch = 20)
legend('topleft', legend = c('Analytic solution: y.true', 'Numerical solution: ODE'),
       col = c('red', 'black'), lty = c(NA, 1), pch = c(20, NA), bty = 'n')
```

### Exploring the Model
```{r}
head(out, 3) #showing the first 3 outputs from the start or 'head' of the data
tail(out, 6) # showing the last 6 outputs from the end or 'tail' of the data

out[, 2] #the first column of the output by column indexing
out[, 'y'] #the first column of the output by column name

out[3, ] #printing the 3rd row
```

# Simulation Using the SIR Model

## Step 1: Code the SIR Model (or function here)
```{r}
SIR = function (t, state, parameters) {
  with(as.list(c(state, parameters)), {
    dS = -beta * S * (I/N); #rate of change of susceptible population
    dI = beta * S * (I/N) - gamma * I; #rate of change of infectious population
    
    dcumI = beta * S * (I/N) #computing the cumulative number of infections
    
    list(c(dS, dI, dcumI)) #returning the rate of change
  })
}
```

## Step 2: Specify Initial Conditions/Parameters
```{r}
N = 1e5 #population size or number of people in the population
I0 = 10 #initial number of infectious peolpe
S0 = N - I0 #intial number of susceptible people
state = c(S = S0, I = I0, cumI = I0) #storing the initial conditions I and S together
parameters = c(beta = 0.5, gamma = 0.3) #storing the model parameters, unit: per day

times = seq(1, 100, by = 1) # setting simulation time steps: 1 to 100 days
``` 

## Step 3: Call the 'ode' function to Generate the Simulation
```{r}
sim = ode(y = state, times = times, func = SIR, parms = parameters)
```


